```{r, echo = FALSE}
library(knitr)
library(ggplot2)
```

```{r setoptions, echo = FALSE}
opts_chunk$set(eval = TRUE)
opts_chunk$set(echo = FALSE)
opts_chunk$set(fig.height = 3.5)
```

The response is the length of teeth in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

```{r}
par(mfrow = c(1,2))

p1 <- ggplot(ToothGrowth, aes(x = factor(dose), y = len, fill = factor(dose))) 
p1 + geom_boxplot() + guides(fill=FALSE) + facet_grid(. ~ supp)

```

After plotting the data as boxplots a number of correlations were observed. The length of the tooth increases as the dosage increases. The OJ delivery method yields a greater length than the VC (approximately 10mm) for smaller dosages but the difference is negligable by a 2mg dosage.


Load the ToothGrowth data and perform some basic exploratory data analyses 

Provide a basic summary of the data.

The tooth growth was compared by supplement and dosage under the null hypothesis that they have equal effect on the tooth.

$H_0: \mu_0 = \mu$







Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose. (Use the techniques from class even if there's other approaches worth considering)
State your conclusions and the assumptions needed for your conclusions.


assumption:
same variance