---
title: "Analysis of US Emissions"
author: "Chris Daly"
date: "Friday, August 22, 2014"
output:
  html_document:
    fig_caption: yes
    theme: spacelab
---
## Synopsis
Fine particulate matter (PM2.5) is an ambient air pollutant for which there is strong evidence that it is harmful to human health. In the United States, the Environmental Protection Agency (EPA) is tasked with setting national ambient air quality standards for fine PM and for tracking the emissions of this pollutant into the atmosphere. Approximatly every 3 years, the EPA releases its database on emissions of PM2.5. This database is known as the National Emissions Inventory (NEI). You can read more information about the NEI at the EPA National Emissions Inventory web site.

For each year and for each type of PM source, the NEI records how many tons of PM2.5 were emitted from that source over the course of the entire year. The data that you will use for this assignment are for 1999, 2002, 2005, and 2008.

The overall goal of this assignment is to explore the National Emissions Inventory database and see what it say about fine particulate matter pollution in the United states over the 10-year period 1999-2008

### Libraries
The following libraries were used throughout the code.

```{r}
library(knitr)
library(ggplot2)
library(plyr)
```

```{r setoptions, echo = FALSE}
opts_chunk$set(cache = TRUE)
```

### Code and Graphs

The data was downloaded from amazon cloudfront, unzipped and loaded into R.
```{r, eval = FALSE}
# check if a data folder exists; if not then create one
if (!file.exists("data")) {dir.create("data")}

# file URL & destination file
fileUrl <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
destfile <- "./data/exdata-data-NEI_data.zip"

# download the file & note the time
download.file(fileUrl, destfile)
dateDownloaded <- date()
```


```{r}
# read in the data until date
NEI <- readRDS("./data/summarySCC_PM25.rds")
SCC <- readRDS("./data/Source_Classification_Code.rds")


# plot 1 - Have total emissions from PM2.5 decreased in the United States 
# 1999-2008?

# aggregate the emissions by year & rename the columns
emissions_by_year <- aggregate(NEI$Emissions, by = list(NEI$year), FUN = sum)

# rename the columns & convert the years to characters
names(emissions_by_year) <- c("Year", "Emissions")
emissions_by_year$Year <- as.character(emissions_by_year$Year)

# adjust the margins
par(mar=c(5.1,5,4.1,2.1))
    
p <- plot(emissions_by_year$Year, emissions_by_year$Emissions/10^6, type = "l", 
     main = expression("Total Emissions from PM"[2.5]*" in the US"),
     xlab = "Year", 
     ylab = (expression("Total" ~ PM[2.5] ~ "Emissions (million tonnes)")))

print(p)

# plot 2 - Have total emissions from PM2.5 decreased in the Baltimore City, 
# Maryland (fips == "24510") from 1999 to 2008?

# subset only the Baltimore City data
BC <- subset(NEI, fips == "24510")

# aggregate the emissions by year & rename the columns
BC_by_year <- aggregate(BC$Emissions, list(BC$year), FUN = sum)

# rename the columns & convert the years to characters
names(BC_by_year) <- c("Year", "Emissions")
BC_by_year$Year <- as.character(BC_by_year$Year)

# adjust the margins
par(mar=c(5.1,5,4.1,2.1))

p <- plot(BC_by_year$Year, BC_by_year$Emissions, type = "l", 
     main = expression("Total Emissions from PM"[2.5]*" in Baltimore City"),
     xlab = "Year",
     ylab = (expression("Total" ~ PM[2.5] ~ "Emissions (tonnes)")))

print(p)

# plot 3 - Of the four types of sources indicated by the type (point, nonpoint, 
# onroad, nonroad) variable, which of these four sources have seen decreases in 
# emissions from 1999-2008 for Baltimore City? Which have seen increases in 
# emissions from 1999-2008? Use the ggplot2 plotting system to make a plot 
# answer this question.

library(ggplot2)

# subset only the Baltimore City data
BC <- subset(NEI, fips == "24510")

# aggregate the emissions by source & year
source_ <- aggregate(BC$Emissions, list(BC$type, BC$year), FUN = sum)

# rename the columns
names(source_) <- c("Type", "Year", "Emissions")

p <- qplot(Year, Emissions, color = Type, data = source_, geom = "path") + 
  ggtitle(expression("Total Emissions from PM"[2.5]*" in Baltimore City by Source")) +
  xlab("Year") +
  ylab(expression("Total" ~ PM[2.5] ~ "Emissions (tonnes)"))

print(p)

# plot 4 - Across the United States, how have emissions from coal 
# combustion-related sources changed from 1999-2008?

# subset the source data based on "Coal" appearing in either sector or name columns
coal_filter <- subset(SCC, grepl("Coal", EI.Sector) | grepl("Coal", Short.Name))

# subset the emissions data based on coal source codes
coal_emissions <- subset(NEI, NEI$SCC %in% coal_filter$SCC)

# aggregate the emissions by year
coal <- aggregate(coal_emissions$Emissions, list(coal_emissions$year), FUN = sum)

# rename the columns
names(coal) <- c("Year", "Emissions")

p <- qplot(Year, Emissions, data = coal, geom = "line") + 
  ggtitle(expression("Total Emissions from PM"[2.5]*" from Coal in the US")) +
  xlab("Year") +
  ylab(expression("Total" ~ PM[2.5] ~ "Emissions (tonnes)"))

print(p)

# plot 5 - How have emissions from motor vehicle sources changed from 1999-2008 
# in Baltimore City?

# subset the source data based on "On-Road" appearing in sector column
motor_filter <- subset(SCC, grepl("On-Road", EI.Sector))

# subset the emissions data based on Balitmore & coal source codes
motor_filter2 <- subset(NEI, fips == "24510" & NEI$SCC %in% motor_filter$SCC)

# aggregate the emissions by year
motor <- aggregate(motor_filter2$Emissions, list(motor_filter2$year), FUN = sum)

# rename the columns
names(motor) <- c("Year", "Emissions")

p <- qplot(Year, Emissions, data = motor, geom = "line") + 
  ggtitle(expression("Total Emissions from PM"[2.5]*" from Vehicles in Baltimore City")) +
  xlab("Year") +
  ylab(expression("Total" ~ PM[2.5] ~ "Emissions (tonnes)")) 

print(p)

# subset the emissions data based on Balitmore or LA & vehicle codes
motor_filter <- subset(NEI, (fips == "24510" | fips == "06037") & type == "ON-ROAD")

# rename the values, for a more interpretable dataframe & graph
motor_filter <- transform(motor_filter, region = ifelse(fips == "24510", "Baltimore City", 
                                                        "LA County"))
# sum the data by year-region (4x2)
motor <- ddply(motor_filter, .(year, region), function(x) sum(x$Emissions))

# rename the columns
names(motor) <- c("Year", "Region", "Emissions")

p <- qplot(Year, Emissions, data = motor, geom = "line", color = Region) + 
  ggtitle(expression("Total Emissions from PM"[2.5]*" from Vehicles in Baltimore City and LA")) +
  xlab("Year") +
  ylab(expression("Total" ~ PM[2.5] ~ "Emissions (tonnes)")) + 
  facet_wrap(~ Region, scales = "free") + theme(legend.position = "none")

print(p)

```
